The `zoon` R package for reproducible and shareable species distribution modelling

### 2016-11-01 ms.Rmd
add future developments text

Tutorials on how to create workflows and modules, as well as full technical details for module developers, are provided as vignettes distributed with `zoon`.

The `zoon` packages provides solutions to some of the technical barriers to the ultimate goal of more open, productive, reproducible SDM research.
Achieving this goal will also depend on overcoming the social hurdle of developing a common tool for comparing and quantitatively evaluating SDMs.

This will require achieving a critical mass of `zoon` modules, so that there is an incentive for species distribution modellers to use the software, and a critical mass of users, so that there is an incentive for SDM methods developers to encode their proposed methods as modules in the community repo.
At this point `zoon` will provide a common environment in which to evaluate SDM methods and enable the SDM research community to make evidence-based decisions about best practice and the goals for development of the field.

In order to facilitate this social aspect, future work will develop an online platform to enable exploring contributed modules and workflows.
This platform will also provide an online space to discuss, and openly evaluate, proposed best practices in SDM.
The zoon R package therefore represents a step towards a more reproducible ecosystem of SDM software.

<!— point to github development site—>

>>>>

Reproducibility, open-data and open-science are topics that receive ever more attention all across science.
Journals (McNutt 2016) and funders (European Commission 2016) are moving towards openly shared data and code resources as a requirement, as was set out in the executive order issued by Barak Obama in 2013 (https://www.whitehouse.gov/the-press-office/2013/05/09/executive-order-making-open-and-machine-readable-new-default-government-).
These trends strongly suggest researchers must at least find ways of documenting and disseminating code and data, if not ways that enable end users. Numerous platforms for publishing analytical outputs exist (e.g. GitHub, FigShare, Dryad, CRAN), but end users will still face the demands of locating, deciphering and managing different resources across different platforms.
Then, the real challenge of open and reproducible science, that has an increased capacity for innovation, is the challenge of increasing the usability of code and data.

However, “scientific software is developed either by domain scientists who have little training in user interface (UI) or experience (UX) design, or by software engineers who build broad systems that do not suit the niche needs or practices of domain scientists” (Thomer et al. 2016).
Then, there is a need for environments that can address niche needs (e.g. of a modelling discipline such as SDM) and that address usability issues for diverse outputs produced by diverse scientists.
The SDM community is large enough to warrant a discipline specific platform such as zoon.
The zoon framework offers a new opportunity for the SDM community to develop community modelling benchmarks, and resolve scientific and methodological questions in ways that the current culture of publishing cannot achieve.
It may be that zoon can also be used for other modelling activities that have similar workflows, or use similar data and processing steps.

During the development of ZOON we have improvised a user-centred, participatory design process that builds on our prior experiences with software development and soliciting user feedback at 3 stages.
An ideation workshop set the initial requirements before any code was written, and two further workshops have allowed users input on an initial prototype distributed via GitHub and then a stable R package.
This is by no means a full UX approach (Pavelin et al. 2012), but it has directly informed the design. 
Continued user engagement will be vital to developing a critical mass of zoon modules that can incentivise modellers to contribute to community resources that enable evidence-based decisions about best practice.
An important step would be the publication of a large model inter-comparison within the zoon environment, such as that produced by Elith et al. (2006) or more recently in Beaumont et al. (2016).
At present, studies such as these might be considered as ‘closed knowledge’ even if they were reproducible from the publication.
In order to facilitate a user –friendly approach to distributed working, future work will include the development of an online interface for accessing a repository of contributed modules and modifying workflows.
This interface might also host methodological discourse and openly evaluate proposed best practices in SDM.
In sum, we hope that the zoon R package represents a step towards a more reproducible ecosystem of SDM software.
